---
author: "Intermediate ggplot: Humanities Visualization in R"
title: Jason A. Heppler
maketitle: true
output:
  rmdshower::shower_presentation:
    self_contained: false
    katex: true
    theme: material
---

```{r, echo=FALSE}
# superfund sites CSV
# superfund_sites <- read.csv("data/superfund_sites.csv") %>%
#   mutate(
#     Latitude = as.numeric(gsub("[^0-9.-]", "", Latitude)), 
#     Longitude = as.numeric(gsub("[^0-9.-]", "", Longitude)) 
#   )  
# sites_keep <- c("St", "Site.Name","Address","City","Zip","County","Latitude","Longitude","Final.Date")
# superfund_sites <- superfund_sites[sites_keep]
# superfund_sites$date <- as.Date(superfund_sites$Final.Date, "%m/%d/%y")
# superfund_sites <- setNames(superfund_sites, tolower(names(superfund_sites)))
# library(lubridate)
# superfund_sites$year <- year(superfund_sites$date)
```

## Advanced ggplot

1.  Theming
2.  Labels and Annotation
3.  Network Visualization
4.  Mapping

## Theming

Themes allow you to control the appearance of **non-data** elements. We can build graphics that are visually appealing through the modification, addition, or subtraction of titles, axis labels, tick marks, and legends. 

Consult the [theming documentation](http://docs.ggplot2.org/current/theme.html) to see all the available options.

---

## Theming

Let's look at a new data set.

```{r, message = F, warning = F}
library(historydata)
data("early_colleges")
early_colleges %>% head(5)
```

## Theming

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  labs(title = "Early Colleges")
```

---

## Theming

`ggplot` comes with a few default themes:

| Theme             | Description                                                         |
| :----             | ------------------------------------------------------------------: |
| theme_bw()        | A theme with white background and black gridlines.                  |
| theme_classic()   | A classic-looking theme, with x and y axis lines and no gridlines.  |
| theme_gray()      | The default theme function.                                         |
| theme_minimal()   | A minimal theme with white background and faint gridlines.          |
| theme_linedraw()  | A theme for making line graphs have high contrast.                  |

## Theming - Minimal

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Early Colleges")
```

## Theming - Classic

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_classic() +
  labs(title = "Early Colleges")
```

## ggthemes

We could also leverage the `ggthemes` package, which collects several theme options you can call upon for your visualizations.

```{r}
# install.packages('ggthemes')
library(ggthemes)
```

## ggthemes - The Economist

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_economist() +
  labs(title = "Early Colleges")
```

## ggthemes - Stephen Few

Using Stephen Few's "[Practical Rules for Using Color in Charts](http://www.perceptualedge.com/articles/visual_business_intelligence/rules_for_using_color.pdf)."

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_few() +
  labs(title = "Early Colleges")
```

## ggthemes - FiveThirtyEight

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_fivethirtyeight() +
  labs(title = "Early Colleges")
```

## ggthemes - Edward Tufte

```{r, fig.height = 4, fig.align = 'center', message = F, warning = F}
ggplot(early_colleges, aes(x = state)) +
  geom_bar() +
  theme_tufte() +
  labs(title = "Early Colleges")
```

## Theming - Creating Your Own

For example:

```{r, eval=F}
theme(
    axis.text = element_text(size = 14),
    legend.key = element_rect(fill = "navy"),
    legend.background = element_rect(fill = "white"),
    legend.position = c(0.14, 0.80),
    panel.grid.major = element_line(colour = "grey40"),
    panel.grid.minor = element_blank(),
    panel.background = element_rect(fill = "navy")
  )
```

## Theming

Let's modify our college counts chart. 

- apply the `ggplot` minimal theme
- make the x-axis labels bold
- rotate the x-asix labels to make them easier to read

## Theming

```{r, fig.height = 3, fig.align = 'center', message = F, warning = F}
library(showtext) # custom fonts
font.add.google("Lato")

ggplot(early_colleges, aes(x = state)) +
  geom_bar(fill = "#007D68") +
  labs(title = "Early Colleges by State") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 15, face = "bold"),
    text = element_text(family="Lato", size=12),
    axis.text = element_text(face = "bold", size = 8),
    axis.text.x = element_text(angle = 90, hjust = 1)
   )
```

## Labels and Annotation

```{r, message = F, warning = F}
data("judges_people")
# Remove rows with missing data
judges_sub <- judges_people[complete.cases(judges_people),]
judges_sub %>% head(5)
```

## Labels and Annotation

```{r, message = F, warning = F, eval = F}
ggplot(judges_sub, 
       aes(x = death_date,
           y = reorder(factor(name_last), death_date))) +
geom_point() +
  labs(x = "", y ="",
       title = "Federal Judges, Death Date")
```

## { .fullpage }

<div class="fullpage width">
```{r, echo=FALSE}
ggplot(judges_sub, 
       aes(x = death_date,
           y = reorder(factor(name_last), death_date))) +
geom_point() +
  labs(x = "", y ="",
       title = "Federal Judges, Death Date")
```
</div>

## Labels and Annotation

```{r, message = F, warning = F, eval = F}
judges.plot.lifespan <- ggplot(judges_sub, aes(x = death_date, 
                            y = reorder(factor(name_last), birth_date)))
judges.plot.lifespan + geom_segment(aes(yend = name_last, xend=birth_date)) + 
  geom_point() +
  theme_minimal() +
  labs( x="Year of Death", 
        y="",
        title="Lifespans of Appointed Judges") +
  theme(panel.grid.major.y=element_blank(), 
        axis.ticks=element_blank(),
        axis.title=element_text(size=8,face="bold"),
        axis.text.y = element_text(size=6)
        )
```

## { .fullpage }

<div class="fullpage width">
```{r, echo=FALSE}
judges.plot.lifespan <- ggplot(judges_sub, aes(x = death_date, 
                            y = reorder(factor(name_last), birth_date)))
judges.plot.lifespan + geom_segment(aes(yend = name_last, xend=birth_date)) + 
  geom_point() +
  theme_minimal() +
  labs( x="Year of Death", 
        y="",
        title="Lifespans of Appointed Judges") +
  theme(panel.grid.major.y=element_blank(), 
        axis.ticks=element_blank(),
        axis.title=element_text(size=8,face="bold"),
        axis.text.y = element_text(size=6)
        )
```
</div>

## { .fullpage }

<div class="fullpage width">
```{r, echo = F}
judges.plot.lifespan + geom_segment(aes(yend = name_last, xend=birth_date, color = death_date < 1980)) + 
  geom_point() +
  theme_minimal() +
  labs( x="Year of Death", 
        y="",
        title="Lifespans of Appointed Judges") +
  theme(panel.grid.major.y=element_blank(), 
        axis.ticks=element_blank(),
        axis.title=element_text(size=8,face="bold"),
        axis.text.y = element_text(size=6)
        )
```
</div>

## Annotation

Let's load up another set of data by reading in a CSV file.

```{r, message = F, warning = F}
# Data prep for next example
superfund <- read.csv("../data/superfund_sites.csv")
superfund.by.state <- superfund %>%
  group_by(St) %>% # St=state
  summarise(count=n())

superfund.by.state %>% head(5)
```

## Annotation

```{r, message = F, warning = F}
ggplot(superfund.by.state, aes(St, count)) +
  geom_bar(stat="identity") +
  labs(title="Superfund Sites by State") +
  theme_minimal() + theme(legend.position='none') +
  labs(x = "State", y = "Count")
```

## Annotation

```{r, fig.height=4, message = F, warning = F}
ggplot(superfund.by.state, aes(St, count)) +
  geom_bar(stat="identity", aes(fill = count > 60)) +
  labs(title="Superfund Sites by State") +
  theme_minimal() + theme(legend.position='none') +
  labs(x = "State", y = "Count") +
  scale_fill_manual(values = c('#4C4C4C', '#8E001C'))
```

## ggrepel

```{r, fig.height=5, message = F, warning = F}
# install.packages('gapminder')
library(gapminder)
keepers <- subset(gap_with_colors,
                continent %in% c("Africa", "Europe") & year == 2007)

ggplot(keepers, aes(x = gdpPercap, y = lifeExp)) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) +
  geom_text(data = subset(keepers, gdpPercap > 30000), aes(label = country))
```

## ggrepel

```{r, fig.height = 5, message = F, warning = F}
# install.packages('gapminder')
# install.packages('ggrepel')
library(gapminder)
library(ggrepel)
keepers <- subset(gap_with_colors,
                continent %in% c("Africa", "Europe") & year == 2007)

ggplot(keepers, aes(x = gdpPercap, y = lifeExp)) + 
  geom_jitter(position = position_jitter(width = 0.1, height = 0), alpha = 1/4) +
  geom_text_repel(data = subset(keepers, gdpPercap > 30000), aes(label = country))
```

## Network Analysis

`ggplot` does not come with a default method for visualizing networks, but communities of users are working to bring network visualization into the grammar of graphics.

Note that these are only for building network visualizations. If you need to do network analysis, you will need to leverage packages like [igraph](http://igraph.org/r/).

## Extending - geom_net

```{r, message = F, warning = F}
# install.packages("geomnet")
library(geomnet)
library(historydata)
data('tudors')

tudors %>% head(5)
```

## Extending - geom_net

```{r, message = F, warning = F}
tudor.network <- ggplot(data = tudors, aes(from_id = person_2, to_id = person_1))
tudor.network + geom_net(label = TRUE, linewidth=0.5, 
           fontsize = 3, size = 6,
           ecolour = ifelse(tudors$relationship == "child", "#a6cee3", "#b2df8a")) +
  theme_net() +
  labs(title = "Tudors")
```

## Extending - geom_net

```{r, message = F, warning = F}
data("lesmis")
lesmisnet <- merge(lesmis$edges, lesmis$vertices, 
                   by.x="from", by.y="label", all=TRUE)
lemisnetwork <- ggplot(data=lesmisnet, aes(from_id=from, to_id=to))
lemisnetwork + geom_net(layout="fruchtermanreingold", 
                        label=TRUE, vjust=-0.5, 
                        aes(linewidth = degree/5)) + theme_net()
```

## Mapping

```{r, message=FALSE, warning=FALSE, eval=F}
library(ggmap)
basemap <- get_map(location="United States", zoom = 4, maptype = "terrain")

ggmap(basemap) +
  geom_point(data=superfund_sites, aes(x=longitude, y=latitude), color="#894D46", size = 0.5, na.rm = TRUE) +
  coord_map() 
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), # get rid of x ticks/text
                 axis.ticks.x = element_blank(),axis.text.x = element_blank(), # get rid of y ticks/text
                 plot.title = element_text(lineheight=.8, face="bold", vjust=1))
```

## { .fullpage }

<div class="fullpage width">
```{r, echo=FALSE}
library(ggmap)
basemap <- get_map(location="United States", zoom = 4, maptype = "terrain")

ggmap(basemap) +
  geom_point(data=superfund_sites, aes(x=longitude, y=latitude), color="#894D46", size = 0.5, na.rm = TRUE) +
  coord_map() +
  theme(axis.ticks.y = element_blank(),axis.text.y = element_blank(), # get rid of x ticks/text
                 axis.ticks.x = element_blank(),axis.text.x = element_blank(), # get rid of y ticks/text
                 plot.title = element_text(lineheight=.8, face="bold", vjust=1))
```
</div>

## Worksheet #2: Intermediate ggplot
